Numerical analysis of the linear vertical shear instability in astrophysical disks

### 2014-12-29 appendix.tex
replace numerical section with that of full equations
extended analytic discussion to thermal relaxation cases
restructing analytic discussion

@@

>>>>

\section{Full governing equation for vertically isothermal disks}\label{adia_improve}
% \section{Improving the nearly-Keplerian approximation for adiabatic
%   disks}\label{adia_improve}
 In \S\ref{analytic_adia} we made the replacement $D\to\Omega_k^2$
 before combining Eq. \ref{adia_iso1}---\ref{adia_iso2} to derive
 Eq. \ref{adia_iso3}.

This procedure ignores the vertical dependence of
 $D=\kappa^2(z) - \sigma^2$.

We show here that this has no significant
 consequence for thin disks.

In the global disk with $\Gamma=1$ we have
\begin{align}\label{dkappa2}
  \frac{\p\kappa^2}{\p z} = 4 \frac{\p\Omega^2}{\p z} + r\frac{\p}{\p
    r}\frac{\p\Omega^2}{\p z} = -
  \frac{\p\ln\rho}{\p z}\frac{qc_s^2}{r^2} \left(2 + q +
    \frac{z^2/r^2-2}{z^2/r^2+1}\right). 
\end{align}
For the local problem we may then write
\begin{align}
  \frac{dD}{dz}  = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2}F(z;q),
\end{align}
where the function $F$ corresponds to the bracket in
Eq. \ref{dkappa2}. This function varies from $F=q$ at $z=0$ to $F\to
3+q$ as $|z|\to\infty$, i.e. its magnitude is of order unity.

 Eliminating $W$ and $Q$ from Eq. \ref{ode_w}---\ref{ode_Q} and
 Eq. \ref{lin_vz} for vertically isothermal disks, retaining the
 vertical dependence of $D$ and including thermal relaxation, we
 obtain 
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\sigma^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z,
\end{align}
where we recall $\chi = \left(1-\ii\sigma t_c\right)/\left(1-\ii\sigma
t_c \gamma\right)$. Making the low-frequency approximation with the
replacement $D\to \Omega_k^2$ gives, in terms of dimensionless
variables,
\begin{align}
   &\delta v_z ^{\prime\prime} + \left[1 + \ii\epsilon q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q \epsilon^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii \epsilon q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}q\epsilon^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\sigma}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3}.
\end{align}

@@

% Differentiating Eq. \ref{adia_iso1} properly gives
% \begin{align}
%   \left(D + \gamma k_x^2 c_s^2\right)\frac{d\Delta}{dz}=& D
%   \frac{d^2\delta v_z}{dz^2} + k_x^2c_s^2\left[\frac{\gamma}{\left(D + \gamma k_x^2 c_s^2\right)}\frac{dD}{dz} + \ii
%     \frac{d\ln\rho}{dz}\left(\ii +
%       \frac{q}{k_xr}\right)\right]\frac{d\delta v_z}{dz}\notag\\
%   &+ \ii k_x^2 c_s^2 \left(\ii +
%       \frac{q}{k_xr}\right)\left[\frac{d^2\ln\rho}{dz^2} -
%       \frac{1}{\left(D + \gamma
%           k_x^2c_s^2\right)}\frac{dD}{dz}\frac{d\ln\rho}{dz}\right]\delta
%     v_z. \label{adia_diso1}
% \end{align}
% The terms that were ignored in deriving Eq. \ref{adia_iso3} are those
% proportional to $dD/dz$ in Eq. \ref{adia_diso1}. Now, in the global
% disk with $\Gamma=1$ we have
% \begin{align}\label{dkappa2}
%   \frac{\p\kappa^2}{\p z} = 4 \frac{\p\Omega^2}{\p z} + r\frac{\p}{\p
%     r}\frac{\p\Omega^2}{\p z} = -
%   \frac{\p\ln\rho}{\p z}\frac{qc_s^2}{r^2} \left(2 + q +
%     \frac{z^2/r^2-2}{z^2/r^2+1}\right). 
% \end{align}
% For the local problem we may then write
% \begin{align}
%   \frac{dD}{dz}  = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2}F(z;q),
% \end{align}
% where the function $F$ corresponds to the bracket in
% Eq. \ref{dkappa2}. This function varies from $F=q$ at $z=0$ to $F\to
% 3+q$ as $|z|\to\infty$, i.e. its magnitude is of order unity. 
% % We estimate the
% % importance of these terms by noting that 
% % \begin{align}
% %   \frac{dD}{dz}\equiv \frac{d\kappa^2}{dz} \simeq \frac{d\Omega^2}{dz}
% %   = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2},
% % \end{align}
% % for thin disks. 
% Eq. \ref{adia_diso1} becomes
% \begin{align}
% \left(D + \gamma k_x^2 c_s^2\right)\frac{d\Delta}{dz}=& D
%   \frac{d^2\delta v_z}{dz^2} - k_x^2c_s^2\frac{d\ln\rho}{dz}\left[ 1 - 
%       \frac{\ii q}{k_xr}  +  \frac{\gamma q c_s^2F}{r^2\left(D +
%         \gamma k_x^2 c_s^2\right)}\right]\frac{d\delta
%     v_z}{dz}\notag\\ 
%   &- k_x^2 c_s^2 \left(1 - 
%       \frac{\ii q}{k_xr}\right)\left[\frac{d^2\ln\rho}{dz^2} + 
%       \frac{qc_s^2F}{r^2\left(D + \gamma
%           k_x^2c_s^2\right)}\left(\frac{d\ln\rho}{dz}\right)^2\right]\delta
%     v_z. \label{adia_diso2}
% \end{align}
% Since $D\sim \Omega_k^2$ in the low-frequency limit, we 
% see from Eq. \ref{adia_diso2} that the neglected terms are
% $O(\epsilon^2)$. We can make this explicit by combining
% Eq. \ref{adia_diso1}---\ref{adia_diso2} and Eq. \ref{adia_iso2}, then
% set $D\to\Omega_k^2$. In terms of non-dimensional variables, the result
% is  
% \begin{align}
%    &\delta v_z ^{\prime\prime} + \left(1 + \ii\epsilon q\hat{k} -
%     \frac{\gamma q \epsilon^2\hat{k}^2F}{1+\gamma
%       \hat{k}^2}\right)\ln\rho^{\prime}\delta v_z^\prime +
%   \left[\left(\frac{1}{\gamma} + \ii \epsilon q
%       \hat{k}\right)\ln\rho^{\prime\prime} - \hat{k}^2\left(1 -
%       \frac{\ii\epsilon
%         q}{\hat{k}}\right)\left(\frac{\gamma-1}{\gamma} +
%       \frac{q\epsilon^2F}{1+\gamma\hat{k}^2}\right)\ln\rho^{\prime
%       2}\right]\delta v_z \notag\\&=
%   -\hat{\sigma}^2\left(\frac{1}{\gamma} + \hat{k}^2\right)\delta v_z\label{adia_diso3}.
% \end{align}   
% Eq. \ref{adia_diso3} differs from Eq. \ref{adia_iso3} by terms
% proportional to $\epsilon^2$. For a thin disk, $\epsilon\ll1$, so
% neglecting these terms has no qualitative effect on the discussion in
% \S\ref{analytic_adia}.

\section{Coefficients for the dispersion relation for perturbations
with thermal relaxation}\label{relax_coeff}
The coefficients of the dispersion relation Eq. \ref{relax_disp} is
given by:
\begin{align}
  &c_0 = M(M+1)\widetilde{A}^2,\\
  &c_1 = \ii\beta\left\{\left(1-\gamma\right)\left[1 +
      \khat^2\left(1+2M\right)^2 - 4 \ii\epsilon q\khat M (M+1)\right] 
    - 2\widetilde{A}^2\gamma M (M+1)\right\},\\
  &c_2 = \left(\khat^2 + 1\right)\widetilde{A} + \beta^2\left\{(1-\gamma)\left[1
      + \gamma \khat^2(1+2M)^2 - 4\ii\epsilon q \khat \gamma M(M+1)
    \right]
    -\gamma^2 \widetilde{A}^2 M(M+1)
  \right\},\\
  &c_3 = \beta\left\{\epsilon q \khat + \gamma \left[\ii + \epsilon q
      k \left(1+2\khat^2\right)\right] - 3\ii - 2\ii
    \khat^2\right\},\\
  &c_4 =
  \beta^2\left(1+\gamma\khat^2\right)\left[\gamma\left(1-\ii\epsilon q
    \khat\right)-2\right] - \left(1+\khat^2\right)^2,\\
&c_5 = 2\ii\beta\left(1+\khat^2\right)\left(1+\gamma\khat^2\right),\\
&c_6 = \beta^2\left(1+\gamma\khat^2\right)^2.
\end{align}

### 2014-12-29 linear_adia.tex
restructing analytic discussion

\subsection{Adiabatic perturbations}\label{analytic_adia}
We now consider general $\gamma$ in the
adiabatic limit ($t_c\to\infty$). This is the opposite regime to
isothermal perturbations. We eliminate variables from 
Eq. \ref{lin_mass}---\ref{lin_energy} in favor of $\delta v_z$ 
and $\Delta \equiv \nabla\cdot\delta\bm{v}$ to obtain
\begin{align}
  &\Delta\left(1 + \frac{\gamma c_s^2 k_x^2}{D}\right) = \frac{d\delta
    v_z}{dz} + \frac{\ii k_x^2}{D}\left(\ii c_s^2 \frac{d\ln\rho}{dz}
    - \frac{r}{k_x}\frac{d\Omega^2}{dz}\right)\delta v_z,\label{adia_iso1}\\
  & -\frac{\sigma^2}{c_s^2}\delta v_z = \gamma \frac{d\Delta}{dz} +
  \frac{d}{dz}\left(\frac{d\ln\rho}{dz}\delta v_z\right) +
  \left(\gamma-1\right)\frac{d\ln\rho}{dz} \Delta.\label{adia_iso2}
\end{align}
%{\bf note: incompressible mode, set delta to zero. only works for
% constant gravity.}

We make the low-frequency approximation and the replacement 
$D\to \Omega_k^2$ in Eq. \ref{adia_iso1}---\ref{adia_iso2}. For
simplicity we do this before combining these equations later. We  
show in Appendix \ref{adia_improve} that making these approximations
after combining  Eq. \ref{adia_iso1}---\ref{adia_iso2} adds unecessary
complexity for thin disks. Here, we also use the fact that in the
global disk, 
\begin{align}
  r\frac{\p \Omega^2}{\p z} = -\frac{\p \ln\rho}{\p z} \frac{qc_s^2}{r},
\end{align}
for the power-law prescription of the midplane sound-speed for
$\Gamma=1$.

Making the above substitutions and eliminating $\Delta$ between Eq. \ref{adia_iso1}---
Eq. \ref{adia_iso2} we obtain, in terms of dimensionless variables defined
previously, an equation for $\delta v_z$, 
\begin{align}
  0 =& \dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime\notag\\
  &+\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right)\right.\notag\\
  &\phantom{0=}\left.-\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}

@@

It is instructive to first ask: how
do the eigenfrequencies and eigenfunctions change when we change
$t_c$ from zero to a small but finite value? For sufficiently small
$t_c$, for which the perturbations may be considered
nearly-isothermal, we expect the new solutions to only differ slighly
from the explicit isothermal solutions developed in \S\ref{iso_poly}.

>>>>

\subsection{Adiabatic perturbations}\label{analytic_adia}
We now consider % general $\gamma$
perturbations in the adiabatic limit ($t_c,\beta\to\infty$). This is the opposite regime to
isothermal perturbations. In this case the parameter $\chi = 1/\gamma$
and Eq. \ref{vertiso_gov_nondim} can be written as 
% We eliminate variables from 
% Eq. \ref{lin_mass}---\ref{lin_energy} in favor of $\delta v_z$ 
% and $\Delta \equiv \nabla\cdot\delta\bm{v}$ to obtain
% \begin{align}
%   &\Delta\left(1 + \frac{\gamma c_s^2 k_x^2}{D}\right) = \frac{d\delta
%     v_z}{dz} + \frac{\ii k_x^2}{D}\left(\ii c_s^2 \frac{d\ln\rho}{dz}
%     - \frac{r}{k_x}\frac{d\Omega^2}{dz}\right)\delta v_z,\label{adia_iso1}\\
%   & -\frac{\sigma^2}{c_s^2}\delta v_z = \gamma \frac{d\Delta}{dz} +
%   \frac{d}{dz}\left(\frac{d\ln\rho}{dz}\delta v_z\right) +
%   \left(\gamma-1\right)\frac{d\ln\rho}{dz} \Delta.\label{adia_iso2}
% \end{align}
% %{\bf note: incompressible mode, set delta to zero. only works for
% % constant gravity.}

% We make the low-frequency approximation and the replacement 
% $D\to \Omega_k^2$ in Eq. \ref{adia_iso1}---\ref{adia_iso2}. For
% simplicity we do this before combining these equations later. We  
% show in Appendix \ref{adia_improve} that making these approximations
% after combining  Eq. \ref{adia_iso1}---\ref{adia_iso2} adds unecessary
% complexity for thin disks. Here, we also use the fact that in the
% global disk, 
% \begin{align}
%   r\frac{\p \Omega^2}{\p z} = -\frac{\p \ln\rho}{\p z} \frac{qc_s^2}{r},
% \end{align}
% for the power-law prescription of the midplane sound-speed for
% $\Gamma=1$.

% Making the above substitutions and eliminating $\Delta$ between Eq. \ref{adia_iso1}---
% Eq. \ref{adia_iso2} we obtain, in terms of dimensionless variables defined
% previously, an equation for $\delta v_z$,

\begin{align}
  0 =& \dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime\notag\\
  &+\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right)\right.\notag\\
  &\phantom{0=}\left.-\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}

@@

It is instructive to first ask the more analytically tractable question: how
do the eigenfrequencies and eigenfunctions change when we change
$t_c$ from zero to a small but finite value? For sufficiently small
$t_c$, for which the perturbations may be considered
nearly-isothermal, we expect the new solutions to only differ slighly
from the explicit isothermal solutions developed in \S\ref{iso_poly}.

### 2014-12-29 linear_iso.tex
restructing analytic discussion

\section{Analytical discussion of vertically isothermal disks} 
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the shearing box. We
examine separately the limiting case of isothermal perturbations
($\gamma\equiv1$ and any value of $t_c$, or any value of $\gamma$ with
$t_c = 0$) and adiabatic perturbations ($\gamma\neq 1$ and $t_c\to\infty$).   
\subsection{Isothermal perturbations}\label{iso_discuss}
The linear problem simplifies considerably for the case
$\Gamma=\gamma=1$.  Then 
\begin{align}
  Q=W. 
\end{align}
Inserting this into Eq. \ref{ode_w} gives a single equation for the linear stability
problem for isothermal perturbations in a vertically isothermal disk, 
\begin{align}\label{iso_ode}
  \frac{d^2W}{dz^2} + \left(\frac{d\ln{\rho}}{dz} - \frac{\ii k_x
      r}{D}\frac{d\Omega^2}{dz}\right) \frac{dW}{dz} +
  \sigma^2\left(\frac{1}{c_s^2} + \frac{k_x^2}{D}\right)W=0, 
\end{align} 
subject to appropriate boundary conditions. Eq. \ref{iso_ode} also
applies to $\gamma>1$ for $t_c\to0$.

We remark that Eq.\ref{iso_ode} is also applicable for    
$\gamma=\Gamma\neq 1$ in the limit $t_c\to\infty$ (i.e. neutrally
stratified adiabatic disks), since in that case  
$Q\simeq W$ as well. However, the following discussion is only
valid for $\gamma=\Gamma=1$ because $c_s$ is taken to be a constant.

\subsubsection{Approximations}
To facilitate an analytical discussion, we simplify the governing
equation Eq. \ref{iso_ode} with the following approximations:

\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. The problem simplifies because the eigenvalue now only
  appears in the last term in Eq. \ref{iso_ode}. We call this the
  low-frequency approximation. 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in Eq. \ref{iso_ode} through
  the factor $d\Omega^2/dz$. We therefore simplify further by ignoring
  the vertical dependence of $\kappa^2$ elsewhere. For clarity, we
  will replace $\kappa^2$ with $\Omega_k^2$. 
  % , although the discussion
  % is unchanged if we replaced 
  % $\kappa^2$ with $\kappa^2(0)$, for example.    
\end{enumerate}
Making both of these approximations is equivalent to the replacement
$D\to\Omega_k^2$.  We will consider sound-speed profiles with $q<0$ in the global 
disk. Then $\Omega^2(z)$ decreases away from the midplane,
and may reach small values for sufficiently large $|z|$, which
invalidates the low-frequency approximation. In principle, then, the
above approximations implicitly limits us to consider thin disks ($\epsilon
\ll 1$) with vertical domain sizes not too large, so that the rotation
is nearly-Keplerian.

\subsubsection{Neccessity of vertical shear for
  instability}\label{integral_relation} 
With the above approximations and introducing the dimensionless variables
\begin{align}
  \hat{z} = z/\Hiso,\quad \hat{k}=k_x\Hiso, \quad \hat{\sigma} = \sigma/\Omega_k,
\end{align}
the governing equation can be re-written as
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \ii \epsilon q \hat{k}
    f(\zhat)\right]W^\prime + \hat{\sigma}^2\left(1+\hat{k}^2\right)W,\label{iso_ode1}
\end{align}
where the primes denote $d/d\zhat$ and $f(\zhat)$ is defined such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = \epsilon^2q f(\hat{z})\Omega_k^2.
\end{align}

We can establish a necessary condition for instability by multipling
Eq. \ref{iso_ode1} by $\rho W^*$ and integrate vertically from
$\zhat=\zhat_1$ to $z=\zhat_2$. We neglect boundary 
terms when integrating by parts, by assuming $W$ or
$W^\prime$ vanishes at the boundaries, or that the boundaries are 
sufficiently far away so that the boundary terms are negligible because of the
decaying background density with increasing height. Then,
\begin{align}
  &\hat{\sigma}^2\left(1+\hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|W|^2d\zhat \notag\\
  &=\int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\zhat 
  +\ii \epsilon q \hat{k}\int_{\zhat_1}^{\zhat_2}\rho f(\zhat) W^*W^\prime d\zhat.\label{integral_relation1}
\end{align}
It follows that for instability ($\imag\hat{\sigma}>0$), it is neccessary to
have $q\neq0$ or more generally $d\Omega^2/dz\neq 0$, i.e. vertical shear.

>>>>

\section{Analytical discussion of vertically isothermal disks} 
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the shearing box. We
examine separately the limiting case of isothermal perturbations
($\gamma\equiv1$ and any value of $t_c$, or any value of $\gamma$ with
$t_c = 0$) and adiabatic perturbations ($\gamma\neq 1$ and
$t_c\to\infty$), before considering the effect of thermal relaxation.

\subsection{Approximate governing equation for general perturbations}\label{approx_gov}
To facilitate a simple discussion, we derive a governing
equation with the following approximations:

\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. % The problem simplifies because the eigenvalue now only
  % appears in the last term in Eq. \ref{iso_ode}.
  We call this the
  low-frequency approximation. 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in the linearized equations through
  the factor $d\Omega/dz$ in Eq. \ref{lin_vy}. We therefore simplify
  further by ignoring the vertical dependence of $\Omega$ and $\kappa$
  elsewhere. For clarity, we replace $\kappa^2$ with $\Omega_k^2$. 
  % , although the discussion
  % is unchanged if we replaced 
  % $\kappa^2$ with $\kappa^2(0)$, for example.    
\end{enumerate}
Making both of these approximations is equivalent to the replacement
$D\to\Omega_k^2$.  After making this replacement, we eliminate $W$ and
$Q$ between  Eq. \ref{lin_vz} and Eq. \ref{ode_w}---\ref{ode_Q},
making use of Eq. \ref{vertical_shear}, to
obtain an equation for $\delta v_z$ when $\Gamma=1$,
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left(1 + \frac{\ii k_x c_s^2
      q}{\Omega_k^2r}\right)\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left[\sigma^2\left(\frac{k_x^2}{\Omega_k^2} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{\Omega_k^2r}\right)\frac{d^2\ln\rho}{dz^2}\right. \notag\\
    &\phantom{+=}\left.-
    \frac{c_s^2}{\Omega_k^2}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left(1-\chi\right) 
   \right]\delta v_z,\label{vertiso_gov}
\end{align}
where
\begin{align}
  \chi \equiv \frac{1-\ii\sigma t_c}{1-\ii\sigma t_c\gamma}.
\end{align}

It is convenient to introduce the dimensionless variables
\begin{align}
  \hat{z} = z/\Hiso,\quad \hat{k}=k_x\Hiso, \quad \hat{\sigma} = \sigma/\Omega_k,
\end{align}
then Eq. \ref{vertiso_gov} becomes
\begin{align}
  0=& \delta v_z ^{\prime\prime} + \left(1 + \ii\epsilon
    q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime \notag\\
  &+
  \left[\hat{\sigma}^2\left(\hat{k}^2+\chi\right) +
    \left(  \chi + \ii \epsilon q\hat{k}\right)\ln\rho^{\prime\prime}\right.
    \notag\\
    &\left.- \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
\end{align}
where $^\prime$ denotes $d/d\zhat$. Eq. \ref{vertiso_gov_nondim} will
be the basis for our discussion of vertically isothermal disks,
although further simplifications or maniputations will be performed
depending the type of perturbations being considered.

In deriving Eq. \ref{vertiso_gov}, the vertical dependence of
$D=\kappa^2(z)-\sigma^2$ was ignored because the replacement
$D\to\Omega_k^2$ was made before eliminating variables in favor of
$\delta v_z$. In Appendix \ref{adia_improve} we show that making this
replacement after eliminating variables, which requires the vertical
derivative of $D$, adds unecessary complexity for thin disks.

We will consider sound-speed profiles with $q<0$ in the global 
disk. Then $\Omega^2(z)$ decreases away from the midplane,
and may reach small values for sufficiently large $|z|$, which
invalidates the low-frequency approximation. In principle, then, the
above approximations implicitly limits us to consider thin disks ($\epsilon
\ll 1$) with vertical domain sizes not too large, so that the rotation
is nearly-Keplerian.

\subsection{Isothermal perturbations}\label{iso_discuss}
Isothermal perturbations corresponds to $\gamma=1$ for any thermal
relaxation timescale $\beta$, or for any adiabatc index $\gamma$ when
$\beta=0$. In this case Eq. \ref{ode_Q} implies 
\begin{align}
  Q = W, 
\end{align}
 and so 
 \begin{align}
   \ii\hat{\sigma}c_s\delta v_z = W^\prime
   % \quad   \text{(isothermal perturbations)},
 \end{align}
from Eq. \ref{lin_vz}. For isothermal perturbations 
the parameter $\chi=1$ so it is easiest to work with
$W$ by integrating Eq. \ref{vertiso_gov_nondim}. For later discussion
it is convenient to write the result as

\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \ii \epsilon q \hat{k}
    f(\zhat)\right]W^\prime + \hat{\sigma}^2\left(1+\hat{k}^2\right)W,\label{iso_ode1}
\end{align}
where $f(\zhat)$ is defined such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = \epsilon^2q f(\hat{z})\Omega_k^2.
\end{align}

% The linear problem simplifies considerably for the case
% $\Gamma=\gamma=1$.  Then 
% \begin{align}
%   Q=W. 
% \end{align}
% Inserting this into Eq. \ref{ode_w} gives a single equation for the linear stability
% problem for isothermal perturbations in a vertically isothermal disk, 
% \begin{align}\label{iso_ode}
%   \frac{d^2W}{dz^2} + \left(\frac{d\ln{\rho}}{dz} - \frac{\ii k_x
%       r}{D}\frac{d\Omega^2}{dz}\right) \frac{dW}{dz} +
%   \sigma^2\left(\frac{1}{c_s^2} + \frac{k_x^2}{D}\right)W=0, 
% \end{align} 
% subject to appropriate boundary conditions. Eq. \ref{iso_ode} also
% applies to $\gamma>1$ for $t_c\to0$.

Note that for isothermal perturbations the issue of ignoring the
vertical dependence of $D$ described in \S\ref{approx_gov} and Appendix
\ref{adia_improve} is irrelevant, since
Eq. \ref{iso_ode1} may be derived directly by substituting $Q=W$ in
Eq. \ref{ode_w}, i.e. no derivatives of $D$ needs to be taken.

We remark that Eq. \ref{iso_ode1} is also applicable for  $\gamma=\Gamma\neq 1$ in the
limit $t_c\to\infty$ (i.e. neutrally stratified adiabatic disks),
since in that case $Q\simeq W$ as well. However, the following
discussion is only  valid for $\gamma=\Gamma=1$ because $c_s$ is taken
to be a constant.

\subsubsection{Neccessity of vertical shear for
  instability}\label{integral_relation} 
We can establish a necessary condition for instability by multipling
Eq. \ref{iso_ode1} by $\rho W^*$ and integrate vertically from
$\zhat=\zhat_1$ to $z=\zhat_2$. We neglect boundary 
terms when integrating by parts, by assuming $W$ or
$W^\prime$ vanishes at the boundaries, or that the boundaries are 
sufficiently far away so that the boundary terms are negligible because of the
decaying background density with increasing height. Then,
\begin{align}
  &\hat{\sigma}^2\left(1+\hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|W|^2d\zhat \notag\\
  &=\int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\zhat 
  +\ii \epsilon q \hat{k}\int_{\zhat_1}^{\zhat_2}\rho f(\zhat) W^*W^\prime d\zhat.\label{integral_relation1}
\end{align}
It follows that for instability ($\imag\hat{\sigma}>0$), it is neccessary to
have $q\neq0$ or more generally $d\Omega^2/dz\neq 0$, i.e. vertical shear.
